Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-24 _tex/index.tex
Built site for gh-pages

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
causal inference framework. Specifically, using the \emph{structural
approach} to causal inference
\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study aims to
construct a scientific model. This model will elucidate the underlying
assumptions of the DCJ system, providing plausible mechanisms for how
the DCJ outcome is generated. Next, using a minimal set of assumptions,
the study will translate the scientific model into a probabilistic
statistical model. This model will produce statistical estimates to draw
inferences about plausible causal relationships within the DCJ system.

@@

\section{Preliminaries}\label{sec-prelim}

\subsection{The structural approach to causal
inference}\label{sec-prelim-structural}

In statistics, \emph{causal inference} refers to the process of
identifying the causes of a phenomenon and estimating their effects
using data \citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical
statistical modeling, which focuses solely on summarizing data and
inferring associations, causal inference provides a coherent
mathematical representation for analyzing causes and counterfactuals
\citep{Pearl_2009}. Counterfactuals reflect scenarios \emph{contrary to
fact}, where alternative \emph{potential} outcomes have not been
observed and cannot be observed \citep{Neal_2020, Counterfactual_2024}.
Using counterfactuals, researchers can construct a theory of the world
that explains why specific causes have specific effects, and what
happens in the absence of those causes. Consequently, counterfactuals
represent the highest level of abstraction in causal inference
\citep{Pearl_et_al_2018}.

Several approaches to constructing counterfactuals and applying causal
inference exist, but two are particularly prominent: the potential
outcomes approach and the structural approach. The potential outcomes
approach, also known as the Neyman-Rubin causal model, was developed by
\citet{Neyman_et_al_1990} and \citet{Rubin_1974}. In contrast, the
structural approach was pioneered by \citet{Wright_1921}, and later
extended by Pearl \citep{Pearl_2009, Pearl_et_al_2016}. Both approaches
employ rigorous mathematical notation to characterize causal inference,
but they do so in different ways \citep{Neal_2020}. The potential
outcomes approach relies on counterfactual notation, whereas the
structural approach utilizes the do-operator and Structural Causal
Models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these
differences, both approaches provide methods for using experimental and
observational data to estimate causal effects \citep{Pearl_2010}.

However, the structural approach offers a significant advantage over the
potential outcomes approach by enabling the graphical representation of
systems through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. These graphical
representations provides a transparent depiction of a system's
complexity, revealing its underlying assumptions and the plausible
mechanisms generating the system's outcome. This capability aligns well
with the aims of this study. Consequently, the structural approach will
be the primary method used to address the study's research goal.

@@

\subsection{Graphs, DAGs and SCMs}\label{sec-prelim-graphs}

\subsection{The flow of association and
causation}\label{sec-prelim-flow}

@@

\textbf{Conflict of interests:} The authors declare no conflict of
interest.

@@

\textbf{Availability of data and materials:} No data was utilized in
this study

@@

\textbf{Counterfactual:}

>>>>

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
causal inference framework. Specifically, using the structural approach
to causal inference \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016},
the study aims to construct a scientific model. This model will
elucidate the underlying assumptions of the DCJ system, providing
plausible mechanisms for how the DCJ outcome is generated. Next, using a
minimal set of assumptions, the study will translate the scientific
model into a probabilistic statistical model. This model will produce
statistical estimates to draw inferences about plausible causal
relationships within the DCJ system.

@@

\section{Theoretical background}\label{sec-background}

\subsection{The structural approach to causal
inference}\label{sec-background-structural}

In statistics, \emph{causal inference} refers to the process of
identifying the causes of a phenomenon and estimating their effects
using data \citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical
statistical modeling, which focuses solely on summarizing data and
inferring associations, causal inference provides a coherent
mathematical representation for analyzing causes and counterfactuals
\citep{Pearl_2009}.

According to \citet{Pearl_et_al_2018}, counterfactuals occupy the
highest level of cognitive abstraction in the ladder of causation,
followed by intervention and association, and form the foundation of
causal inference. Counterfactuals represent scenarios \emph{contrary to
fact}, where alternative \emph{potential} outcomes resulting from a
cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. Nevertheless, despite their
abstract nature, counterfactuals enable the development of a
\emph{theory of the world} that explains why specific causes have
specific effects and what occurs in the absence of those causes
\citep{Pearl_et_al_2018}. Counterfactuals achieves this by translating
causal statements to statements about ``what would have happened under
different circumstances.''

Several approaches to causal inference and counterfactuals exist, but
two are particularly prominent: the potential outcomes approach, also
known as the Neyman-Rubin causal model
\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}. Both approaches
employ rigorous mathematical notation to characterize causal inference,
but they do so in different ways \citep{Neal_2020}. The potential
outcomes approach relies on counterfactual notation, whereas the
structural approach utilizes the do-operator and structural causal
models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these
differences, both notations can be expressed in terms of the other, and
both approaches provide methods for using experimental and observational
data to estimate causal effects \citep{Pearl_2010}.

However, the structural approach offers a key advantage over the
potential outcomes approach: it enables the graphical representation of
systems through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. These graphical
representations provide a transparent depiction of a system's
complexity, revealing its underlying assumptions and the plausible
causal mechanisms responsible for generating the system's outcome.
Ultimately, DAGs play the important role of effectively conveying the
assumed causal structure of the system, and also determine wich
variables should be controlled for to be able to estimate causal
effects. Consequently, the structural approach will be the primary
method used to address one of the study's research goals.

\subsection{DAGs, SCMs, and the flow of association and
causation}\label{sec-background-dag}

As previously mentioned, the structural approach employs Directed
Acyclic Graphs (DAGs) for the graphical representation of systems. A
\emph{graph} consists of nodes connected by edges
\citep{Gross_et_al_2018}. In a \emph{directed graph}, edges extend from
a \emph{parent} node to a \emph{child} node, with arrows indicating the
direction of causal influence. In directed \emph{acyclic} graphs, the
direction of causal influences does not loop back on themselves, meaning
the graph contains no cycles \citep[@McElreath\_2020]{Neal_2020}.

\subsection{But where does it all fit?}\label{sec-background-where}

@@

@@

\textbf{Financial interests:} The authors have no relevant financial
interest to disclose.

\textbf{Non-financial interests:} Author XX serve on advisory broad of
Company Y but receives no compensation this role.

@@

\textbf{Availability of data and materials:} No data was utilized in
this study.

@@
